---
title: "Simulating Drop-seq scRNA count data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
# Install scSimu and related packages
library(fastmap)

if (!requireNamespace("scSimu", quietly = TRUE)){
  devtools::install_github("xs222/scSimu")
}
if (!requireNamespace("glmGamPoi", quietly = TRUE)){
  if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
  BiocManager::install("glmGamPoi")
}
if (!requireNamespace("CSCORE", quietly = TRUE)){
  devtools::install_github("ChangSuBiostats/CS-CORE") 
}

if (!requireNamespace("SeuratData", quietly = TRUE)){
  devtools::install_github('satijalab/seurat-data')
}

# load package
library(scSimu)
```

# Simulating droplet-based data

## Load data

Here we utilize the T cells from 2,700 PBMC from Seurat.

```{r}
SeuratData::InstallData("pbmc3k")
library(Seurat)
library(SeuratData)
library(ggplot2)
library(patchwork)

pbmc3k.final <- LoadData("pbmc3k", type = "pbmc3k.final")
pbmc3k.final <- UpdateSeuratObject(pbmc3k.final)
pbmc3k.final
table(pbmc3k.final$seurat_annotations)
```

```{r}
pbmc3k.T <- subset(pbmc3k.final, subset = seurat_annotations %in% c("Naive CD4 T", "Memory CD4 T", "CD8 T"))
pbmc3k.T
```

## Estimate marginal parameters

```{r}
pbmc3k_count <- as.matrix(GetAssayData(object = pbmc3k.T, slot = "counts"))
marginal_para <- marginal_fit(pbmc3k_count)
```

Check the relationship between mu and alpha.

```{r}
ggplot(marginal_para, aes(x=mu, y=alpha))+
  geom_point()+labs(title = "log10(alpha) v.s. log10(mu)")+
  xlab("log10(mu)")+ylab("log10(alpha)")
```

Fit a smooth curve between mu and alpha using the major cluster

```{r}
marginal_para$up <- ifelse(marginal_para$alpha>3.5, "upper","lower")
ggplot(marginal_para, aes(x=mu, y=alpha, color=up))+
  geom_point()+labs(title = "log10(alpha) v.s. log10(mu)")+
  xlab("log10(mu)")+ylab("log10(alpha)")

## fit line (only use the major cluster)
marginal_para_sel <- marginal_para[marginal_para$up=="lower",]
## kernel smooth
km5 <- ksmooth(marginal_para_sel$mu, marginal_para_sel$alpha,
               kernel="normal", bandwidth = bw.SJ(marginal_para_sel$mu)*5)

ksmooth_df <- data.frame(mu = km5$x, alpha = km5$y)
# Plot the data and the fitted line
ggplot() +
  geom_point(data = marginal_para, aes(x = mu, y = alpha), alpha = 0.6) +
  geom_line(data = ksmooth_df, aes(x = mu, y = alpha), color = "red", size = 1) +
  labs(title = "Scatter Plot with Fitted Line",
       x = "log10(mu)",
       y = "log10(alpha)") 
```

Use the fitted curve to find alpha.

```{r}
log10mu <- marginal_para$mu
gene_name <- marginal_para$gene
names(log10mu) <- gene_name
mu <- 10^log10mu

fitted_trend <- data.frame(mu=km5$x, alpha=km5$y)
log10alpha <- rep(NA,nrow(marginal_para))
names(log10alpha) <- marginal_para$gene
for (i in 1:nrow(marginal_para)){
  idx <- which.min(abs(log10mu[i]-fitted_trend$mu))
  log10alpha[i] <- fitted_trend$alpha[idx]
}
alpha <- 10^log10alpha
```

## Simulate independent data

```{r}
simu_ind <- scSimu(mu, alpha, pbmc3k_count)
```

## Simulate correlated data

Consider the correlation structure for highly expressed genes.

```{r}
marginal_para <- marginal_para[order(marginal_para$mu, decreasing = T),]
cor_gene <- marginal_para$gene[1:500]
simu_cor <- scSimu(mu, alpha, pbmc3k_count, IND = F, cor_gene = cor_gene)
```

# Simulating Smart-Seq data

```{r}
# Install related packages
if (!requireNamespace("biomaRt", quietly = TRUE)){
  if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
  BiocManager::install("biomaRt")
}
if (!requireNamespace("Matrix", quietly = TRUE)){
  install.packages("Matrix")
}

if (!requireNamespace("MASS", quietly = TRUE)){
  install.packages("MASS")
}
if (!requireNamespace("pscl", quietly = TRUE)){
  install.packages("pscl")
}

```

## Load data

We use Smart-Seq data from <https://www.nature.com/articles/s41587-020-0465-8>.

```{r}
library(Matrix)
library(dplyr)
library(biomaRt)
library(glmGamPoi)
```

```{r}
count_data = readMM(paste0("C:/Users/15177/Downloads/counts.read.txt"))
cells = read.table(paste0("C:/Users/15177/Downloads/cells.read.new.txt"))
genes = read.table(paste0("C:/Users/15177/Downloads/genes.read.txt"))
meta = read.table(paste0("C:/Users/15177/Downloads/meta.txt"), sep = "\t", header=T)
colnames(count_data) = cells$V1
rownames(count_data) = genes$V1
meta_sub = meta[meta$Method=="Smart-seq2" & meta$Experiment=="pbmc1",]

ct = "Cytotoxic T cell"
cell_name = meta_sub$NAME[meta_sub$CellType==ct]
count_sub = count_data[,cell_name]
count_sub = as.matrix(count_sub)

# filter genes
gene_idx = which((rowSums(count_sub)>=10) & (rowSums(count_sub>=1)>=10))
nrow(count_sub)-length(gene_idx)
ct_counts = count_sub[gene_idx,]
dim(ct_counts)
```

## Identify genes with zero inflation

This is based on the procedures in scDesign2.

```{r, message = FALSE, warning = FALSE, error = FALSE}
check_marginal_fit = function(gene){
    library(pscl)
    library(MASS)
    m <- mean(gene)
    v <- var(gene)
    if(m >= v){
        mle_Poisson <- glm(gene ~ 1, family = poisson)
        res = tryCatch({
          mle_ZIP <- zeroinfl(gene ~ 1|1, dist = 'poisson')
          chisq_val <- 2 * (logLik(mle_ZIP) - logLik(mle_Poisson))
          pvalue = as.numeric(1 - pchisq(chisq_val, 1))
          c(pvalue,0, ifelse(pvalue<0.05, "ZIP", "Poisson"))
        }, error = function(cond){c(1,0, "Poisson")})
    }else{
        res = tryCatch({
                mle_NB <- glm.nb(gene ~ 1)
                tryCatch({
                    mle_ZINB <- zeroinfl(gene ~ 1|1, dist = 'negbin')
                    chisq_val <- 2 * (logLik(mle_ZINB) - logLik(mle_NB))
                    pvalue = as.numeric(1 - pchisq(chisq_val, 1))
                    c(0, pvalue, ifelse(pvalue<0.05, "ZINB", "NB"))
                }, error = function(cond){c(0, 1, "NB")})
             }, warning = function(w) {
                 tryCatch({
                    mle_ZINB <- zeroinfl(gene ~ 1|1, dist = 'negbin')
                    chisq_val <- 2 * (logLik(mle_ZINB) - logLik(mle_NB))
                    pvalue = as.numeric(1 - pchisq(chisq_val, 1))
                    c(0, pvalue, ifelse(pvalue<0.05, "ZINB", "NB"))
                }, error = function(cond){c(0, 1, "NB")})
             })
    }
    return(res)
}

p_gene = list()
for (i in rownames(ct_counts)){
    gene = ct_counts[i,]
    p_gene[[i]] = check_marginal_fit(gene)
}
p = do.call(rbind, p_gene)
colnames(p) = c("Poisson", "NB","group")
p = as.data.frame(p)
p$gene = rownames(p)

ZINB_genes = p$gene[p$group %in% c("ZINB", "ZIP")]
other_genes = setdiff(p$gene, ZINB_genes)

```

## Estimate marginal parameters

we use a Gamma-Poisson GLM for non-zero-inflated genes and a zero-inflated model for zero-inflated genes.

```{r}
## estimate non-zero inflated genes
gp_ex <- glm_gp(ct_counts[other_genes,], size_factors = colSums(ct_counts), verbose = T, overdispersion_shrinkage = T, do_cox_reid_adjustment = T)
vanilla <- data.frame(mu = exp(gp_ex$Beta[,1]), alpha = 1/gp_ex$overdispersions, gene=other_genes)

## estimate zero inflated genes
zinb_res = as.data.frame(matrix(nrow=length(ZINB_genes), ncol=4))
colnames(zinb_res) = c("gene", "zero_prop", "alpha", "mu")
rownames(zinb_res) = ZINB_genes
for (g in ZINB_genes){
    df <- data.frame(counts = ct_counts[g,], library_size = log(colSums(ct_counts)))
    fit_ZINB <- zeroinfl(counts ~ 1+offset(library_size)|1, dist = 'negbin', data=df)
    zinb_res[g,] = c(g, plogis(fit_ZINB$coefficients$zero), fit_ZINB$theta, exp(fit_ZINB$coefficients$count))
}
zinb_res[,2:4] = apply(zinb_res[,2:4], 2, as.numeric)
zinb_res$outlier = ifelse(log10(zinb_res$alpha)>(quantile(log10(zinb_res$alpha), 0.75)+2*IQR(log10(zinb_res$alpha))), 1, 0)

```

```{r}
# fit the kernel regression
vanilla$up <- ifelse(log10(vanilla$alpha)>2.5, "upper","lower")
marginal_fit_PNAS_sel <- vanilla[vanilla$up=="lower",]
km5 <- ksmooth(log10(marginal_fit_PNAS_sel$mu), log10(marginal_fit_PNAS_sel$alpha),
               kernel="normal", bandwidth = bw.SJ(log10(marginal_fit_PNAS_sel$mu))*5)

mu <- vanilla$mu
gene_name <- vanilla$gene
names(mu) <- gene_name
log10mu = log10(mu)

fitted_trend <- data.frame(mu=km5$x, alpha=km5$y)
log10alpha <- rep(NA,nrow(vanilla))
names(log10alpha) <- vanilla$gene
for (i in 1:nrow(vanilla)){
  idx <- which.min(abs(log10mu[i]-fitted_trend$mu))
  log10alpha[i] <- fitted_trend$alpha[idx]
}
alpha <- 10^log10alpha


marginal_fit_zinb <- zinb_res[zinb_res$outlier==0,]
km5_zinb <- ksmooth(log10(marginal_fit_zinb$mu), log10(marginal_fit_zinb$alpha),
               kernel="normal", bandwidth = bw.SJ(log10(marginal_fit_zinb$mu))*5)

mu_zinb <- zinb_res$mu
gene_name_zinb <- zinb_res$gene
names(mu_zinb) <- gene_name_zinb
log10mu_zinb = log10(mu_zinb)

fitted_trend_zinb <- data.frame(mu=km5_zinb$x, alpha=km5_zinb$y)
log10alpha_zinb <- rep(NA,nrow(zinb_res))
names(log10alpha_zinb) <- zinb_res$gene
for (i in 1:nrow(zinb_res)){
  idx <- which.min(abs(log10mu_zinb[i]-fitted_trend_zinb$mu))
  log10alpha_zinb[i] <- fitted_trend_zinb$alpha[idx]
}
alpha_zinb <- 10^log10alpha_zinb

alpha_comb = c(alpha, alpha_zinb)
mu_comb = c(mu, mu_zinb)
```

```{r}
zero_inflated_gene = zinb_res$gene
zero_inflated_p = zinb_res$zero_prop
```

## Estimate correlation structure

For best results, we recommend that users provide their own pre-estimated correlation structure. This is an alternative to the function's default method (CS-CORE), which estimates the structure using a negative binomial model.

```{r}
# generate gene cor mat
genes = rownames(ct_counts)
ensg_ids <- sapply(strsplit(genes, "_"), function(x) x[1])

ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
gene_lengths <- getBM(attributes = c('ensembl_gene_id', 'transcript_length'),
                      filters = 'ensembl_gene_id',
                      values = ensg_ids,
                      mart = ensembl)

gene_info <- gene_lengths %>%
  group_by(ensembl_gene_id) %>%
  summarise(length = mean(transcript_length))

calculate_tpm <- function(counts, lengths) {
  # Ensure lengths are in kilobases
  lengths_kb <- lengths / 1000
  
  # 1. Normalize for gene length
  rpk <- counts / lengths_kb
  # 2. Normalize for sequencing depth
  per_million_scaling_factor <- colSums(rpk) / 1e6
  tpm <- sweep(rpk, 2, per_million_scaling_factor, "/")
  return(tpm)
}

# mean expression level under each cell type
sel_genes = sapply(strsplit(rownames(ct_counts), "_"), function(x) x[1])
matched_lengths <- gene_info$length[match(sel_genes, gene_info$ensembl_gene_id)]
# Calculate TPM, removing genes with no length information
valid_genes <- !is.na(matched_lengths)
tpm_matrix <- calculate_tpm(ct_counts[valid_genes, ], matched_lengths[valid_genes])
log_tpm_matrix <- log1p(tpm_matrix)
log_tpm_matrix = as.matrix(log_tpm_matrix)

mean_exp = rowMeans(log_tpm_matrix)
mean_exp = mean_exp[order(mean_exp, decreasing=T)]
smart_high_exp = names(mean_exp)[1:1000]
cor_mat = cor(t(log_tpm_matrix[smart_high_exp,]))
```

## Simulate independent data

```{r}
smart_simu = scSimu_ZINB(mu_comb, alpha_comb, zero_inflated_gene, zero_inflated_p,
                        count_dat=ct_counts, IND=T, seed=7232024)
```

## Simulate correlated data

```{r}
smart_simu = scSimu_ZINB(mu_comb, alpha_comb, zero_inflated_gene, zero_inflated_p,
                        count_dat=ct_counts, IND=F, cor_mat=cor_mat, seed=7232024)
```
